# Comparing CAMDA metadata related with NCBI

Aim: Find how many species are correctly annotated in both databases CAMDA and NCBI related with CAMDA information (training and test datasets).

-   **Author**: Evelia Coss
-   **Date**: 12/June/2025

## Load packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DT)        # Tablas bonitas
```

## STEP 1. Import Metadata and change format

```{r importData}
# getwd() # [1] "C:/Users/ecoss/Documents/Respaldos/Github_notes/CAMDA2025_AntibioticResistance/scripts"
getwd()

# > training dataset
training_metadata <- read_csv("../rawdata/TrainAndTest_dataset/training_dataset.csv")
# rename species
training_metadata_db <- training_metadata %>%
  mutate(scientific_name_CAMDA = paste(genus, species, sep = " ")) 
# Rows: 6144 Columns: 16── Column

# > test dataset
test_metadata <- read_csv("../rawdata/TrainAndTest_dataset/testing_template.csv")
# rename species
test_metadata_db <- test_metadata %>%
  mutate(scientific_name_CAMDA = paste(genus, species, sep = " ")) 
# Rows: 5345 Columns: 6── Column

# > Load public metadata from NCBI
sra_metadata <- read_csv("../metadata/sra-metadata.csv") 
sra_metadata_db <- sra_metadata %>% janitor::clean_names() %>%  # para limpiar y estandarizar los nombres de columnas de un data frame
  select(run, library_strategy, library_selection, library_source, platform, model, 
                        bio_project, bio_sample, scientific_name) %>%
  rename(accession = run, scientific_name_complete = scientific_name)
# Rows: 11510 Columns: 47── Column specification

# Reduce name
# Quedarme con las primeras dos palabras
sra_metadata_db$scientific_name_NCBI <- word(sra_metadata_db$scientific_name_complete, 1, 2)
```

## STEP 2. Cleaning metadata information

The information for the SRA IDs in the test and training datasets was verified:

1.  SRA IDs found in both test and training sets were removed from the test set, keeping them only in the training set.
2.  Duplicate SRA IDs in the rows of the training dataset were removed.
3.  SRA IDs with all three phenotypes (S/I/R) were removed from the training dataset.
4.  Re-assignment of SRA IDs with duplicated annotations or conflicting annotations.

### 2.1 Exclude overlapping SRA IDs from the test set, retaining them only in the training set

```{r}
# Obtener los IDs que están en ambos dataframes
sraids_overlapping <- intersect(test_metadata$accession, training_metadata$accession)

# Mostrar los IDs comunes
length(sraids_overlapping)
```

We detected `length(sraids_overlapping)` SRA IDs se encuentran compartidos entre ambos archivos.

### Eliminar los IDs que se comparten con training

```{r}
testing_metadata_cleaned <- test_metadata %>% filter(!(accession %in% training_metadata$accession))

# 4055 SRA IDs accessions by row
nrow(testing_metadata_cleaned) 
# check: length(unique(testing_metadata_cleaned$accession))

# Save file
# Create new folder
#dir.create("../rawdata/TrainAndTest_cleaned")
write_tsv(testing_metadata_cleaned, file = "../rawdata/TrainAndTest_cleaned/testing_metadata_cleaned.tsv", quote = "none")
```

## STEP 3. Remove duplicated SRA IDs from the training set

### Cuantos SRA IDs se encuentran repetidos en training

```{r}
# How many unique that we have?
cat("Number of SRA ID unique (accessions)", length(unique(training_metadata$accession)), sep = " ") # 5458 SRA ID uniqued

# Numero de SRA IDs que se repiten varias veces?
training_metadata_duplicated  <- table(training_metadata$accession)
sum(training_metadata_duplicated > 1) # 502 SRA IDs

# How many duplicated that we have?
training_duplicated_table <- training_metadata %>%
  group_by(accession) %>%
  filter(n() > 1) %>%   # filtra accession con más de una ocurrencia
  summarise(duplicados = n()) %>%
  arrange(desc(duplicados))

# Only IDs (502 SRA IDs)
training_SRAIDs_duplicated <- training_duplicated_table$accession
```

```{r}
training_duplicated_db <- training_metadata %>% 
  filter(accession %in% training_SRAIDs_duplicated)
```

## Check: Do we have the same species?

### Training vs NCBI data

```{r}
intersect(training_metadata_db$scientific_name_CAMDA, sra_metadata_db$scientific_name_NCBI)
```

### Test vs NCBI data

```{r}
intersect(test_metadata_db$scientific_name_CAMDA, sra_metadata_db$scientific_name_NCBI)
```

## STEP 4. Comparing the information of Specie and accession (SRA ID)

### Training vs NCBI data

#### 2.1 Check species name matches between CAMDA and NCBI

Find how many species have matching names between CAMDA and NCBI to identify correctly annotated entries.

```{r}
training_completeInfo_db <- left_join(training_metadata_db, sra_metadata_db, by ="accession") %>%
  select(accession, scientific_name_CAMDA, scientific_name_NCBI) %>%
  distinct() %>% #eliminas filas
  # Agregar informacion si coinden o no las especies
  mutate( status = case_when(
      scientific_name_CAMDA == scientific_name_NCBI ~ "coincide",
      TRUE ~ "no coincide"
    )) 

# Global information
training_completeInfo_db %>%  
  # Visualizar informacion en una tabla bonita
  DT::datatable()
```

Save data

```{r save training_completeInfo_db}
write_tsv(training_completeInfo_db, file = "../rawdata/Training_status_sp_CAMDAvsNCBI.tsv.gz")
```

#### 2.2 How many SRA entries are correctly annotated and how many are not?

```{r}
# Global 
table(training_completeInfo_db$status)
```

How many species have at least one misannotated entry?

```{r}
training_completeInfo_db %>% filter(status == "no coincide") %>%
  distinct(scientific_name_CAMDA)
```

#### 2.3 What information recorded in CAMDA does not match the NCBI data?

```{r}
training_completeInfo_db %>% filter(status == "no coincide") %>%
  group_by(scientific_name_CAMDA) %>%
  distinct(scientific_name_NCBI) %>%
  # Visualizar informacion en una tabla bonita
  DT::datatable()
```

#### 2.4 How many samples from each species do not match the NCBI data?

```{r}
training_completeInfo_db %>%
  filter(status == "no coincide") %>%
  group_by(scientific_name_CAMDA) %>%
  summarise(n_mal_anotadas = n()) %>%
  arrange(desc(n_mal_anotadas))
```

## Test vs NCBI data

#### 2.1 Check species name matches between CAMDA and NCBI

```{r}
test_completeInfo_db <- left_join(test_metadata_db, sra_metadata_db, by ="accession") %>%
  select(accession, scientific_name_CAMDA, scientific_name_NCBI) %>%
  distinct() %>% #eliminas filas
  # Agregar informacion si coinden o no las especies
  mutate( status = case_when(
      scientific_name_CAMDA == scientific_name_NCBI ~ "coincide",
      TRUE ~ "no coincide"
    )) 

# Global information
test_completeInfo_db %>%  
  # Visualizar informacion en una tabla bonita
  DT::datatable()
```

Save data

```{r}
write_tsv(test_completeInfo_db, file = "../rawdata/Test_status_sp_CAMDAvsNCBI.tsv.gz")
```

#### 2.2 How many SRA entries are correctly annotated and how many are not?

```{r}
# Global 
table(test_completeInfo_db$status)
```

Que especies no coinden

```{r}
test_completeInfo_db %>% filter(status == "no coincide") %>%
  distinct(scientific_name_CAMDA)
```

#### 2.3 What information recorded in CAMDA does not match the NCBI data?

```{r}
test_completeInfo_db %>% filter(status == "no coincide") %>%
  group_by(scientific_name_CAMDA) %>%
  distinct(scientific_name_NCBI) %>%
  # Visualizar informacion en una tabla bonita
  DT::datatable()
```

#### 2.4 How many samples from each species do not match the NCBI data?

```{r}
test_completeInfo_db %>%
  filter(status == "no coincide") %>%
  group_by(scientific_name_CAMDA) %>%
  summarise(n_mal_anotadas = n()) %>%
  arrange(desc(n_mal_anotadas))
```

## To do list

-   Cuantos SRR tenemos en total por especie?
-   Cuantas 341 se pueden reeetiquetar con la misma especie relacionada con las especies que tenemos?
-   Porcentajes de datos que debemos eliminar?
